---
title: "Student Loan Forecasting"
author: "Frank Laudert"
date: "2023-02-26"
output:
  html_document:
    theme: cerulean
    toc: true
    toc_float: true
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```




# Introduction


<br/><br/>

The purpose of the time series analysis is to forecast total loan
disbursements received at the beginning of each academic term. The
federal direct loan data set includes daily transactions of loan
disbursements from 2015 through 2020.

The R programming language was use for this analysis.

<br/><br/>


```{r}





library(dplyr)
library(lubridate)
library(plotly)
library(ggplot2)
library(stats)
library(xts)
library(zoo)
library(TSstudio)
library(forecast)
library(plotly)
library(gt)
library(gtsummary)
library(tidyverse)
library(modelsummary)
library(summarytools)
library(Hmisc)
library(xtable)
library(stargazer)
library(pander)







```


```{r include=FALSE}


dl_loans<-readRDS("dl_loans")



```




```{r}




knitr::kable(head(dl_loans), "simple")


```



```{r}






describe(dl_loans)


```



```{r}



pander(summary(dl_loans))



```




 view structure of data frame

```{r}



str(dl_loans)




```


# Data Preparation

<br/><br/>

There are three terms that span twelve months of the academic calendar.
The academic calendar and the term start months are displayed below.

<br/><br/>


```{r}



college_calendar <- tibble::tribble(
       ~Month,    ~Term,
  "September",   "Fall",
    "October",   "Fall",
   "November",   "Fall",
   "December",   "Fall",
    "January", "Spring",
   "Feburary", "Spring",
      "March", "Spring",
      "April", "Spring",
        "May", "Spring",
       "June", "Summer",
       "July", "Summer",
     "August", "Summer"
  )

require(knitr)
require(kableExtra)
kable_styling(
              kable(college_calendar, digits = 3, row.names = FALSE, align = "c",
              caption = NULL, format = "html"),
        bootstrap_options = c("striped", "hover", "condensed"),
        position = "center", full_width = FALSE) 






```


<br/><br/>


```{r}


college_calendar %>% 
  kbl(caption="Table 1 Academic Calendar") %>%
  kable_classic("striped", full_width = T) %>%
  row_spec(c(1,5,10), bold = T, color = "white", background = "#3399FF")




```



<br/><br/>

## Aggregate Data


The daily transactions will be aggregated monthly to prepare the the
data for forecasting.


<br/><br/>




```{r}




dl_month<-dl_loans  %>% 
  group_by(month=floor_date(To_Bus_Disb_Date, "month")) %>%
  summarise(disbursed=sum(net_amt))



```


<br/><br/>


```{r}




pander(summary(dl_month),caption= "Table 2 Data Summary")



```


<br/><br/>

The summary of the monthly aggregated data shows 9/1/2015 as the start
date and 12/1/2020 as the end date. The data will be filtered so the
beginning date is 1/1/2016. The summary also shows that there is one
missing value (NA). This missing value will have to be removed before a
forecast can be generated.

<br/><br/>




```{r}




dl_month_2<-dl_month %>% 
  filter(month > '2015-12-31')



```

<br/><br/>

remove NA's from data set

<br/><br/>


```{r}




dl_month_2<-na.omit(dl_month_2)


```


<br/><br/>


```{r}




pander(summary(dl_month_2), caption="Table 3 Monthly Data Summary")


```




<br/><br/>

The summary of the monthly data frame in *Table 3* now has 1/1/2016 as
the start date and the missing value has been removed.

<br/><br/>


dispaly first twelve months of the data set.

<br/><br/>

```{r fig.cap="Table 4 First Twelve Months of dl_month_2"}




head(dl_month_2, n=12) %>%
  kbl(caption="Table 4: First Twelve Rows of Monthly Data") %>%
  kable_classic(full_width=F, html_font = "Cambria")





```


<br/><br/>

Viewing the first twelve rows of the monthly data frame from *Table 4*
we can see that the data is now grouped into months.

<br/><br/>

## Time Series Object

<br/><br/>

The data now will be transformed into a ts (time series) object. This
step prepares the data for forecasting.

<br/><br/>

Select the disbursed column from dl_month

```{r}


dl<-dl_month_2[,2]


```

Convert the dl data frame into a time series object

```{r}

ts_dl<-ts(dl, start=c(2016,1), end=c(2020,12), frequency = 12)




```

<br/><br/>

```{r}


pander(ts_info(ts_dl), caption="Table 5: Time Series Object")



```

<br/><br/>



```{r}







```



```{r}







```



```{r}






```



```{r}







```








